\textbf{Empirical Bayes shrinkage, and denoising of Poisson and heteroskedastic Gaussian signals (?)}

### 2016-02-09 paper/draft_5.tex
updated section on gaussian eg

Talk about this being a fast one-step procedure. Compare with Delouille and note that we do not have to manually remove detail coefficients. Could also do better visually by using better interpolation methods
\newpage
\subsection{ChIP-Seq Data}
Here we apply our Poisson de-noising procedure to next generation sequencing data, commonly seen in the field of genomics. Specifically, we chose an example dataset from the ENCODE (\textbf{Enc}yclopedia \textbf{O}f \textbf{D}NA \textbf{E}lements) project launched by the National Human Genome Research Institute. This dataset contains reads from chromatin immunoprecipitation sequencing (ChIP-seq) measuring transcription factor binding in two different cell types, with two samples for each cell type. Due to the massive size of the data, we selected a representative portion of the reads of length $2^{15}$ from chromosome 1.
%Since we are looking mainly at univariate denoising here, we ran our method on each sample separately to produce the the plots in Figure \ref{fig:seq_smooth}.\\
%\begin{figure}[h]
%\centering
%\includegraphics[scale=0.7]{smoothing_all.pdf}
%\caption{Estimated intensities from SMASH for 4 samples; the first two samples are two replicates from GM12878, a lymphoblastoid cell line, and the last two samples are two replicates from H1hesc, which are human embryonic stem cells. The reads are taken from position 880001 to 1011072 on chromosome 1.}
%\label{fig:seq_smooth}
%\end{figure}
%\bigskip
One goal of analyzing ChIP-seq data is to discover regions where transcription factors are likely to bind to DNA, thereby allowing us to better understand the mechanisms underlying gene regulation. These binding regions are often reflected in the data as ``peaks'', where more counts are present than background noise. Our method allows us to identify these peaks by looking at the estimated intensity function.
%In Figure \ref{fig:seq_smooth}, we immediately notice the presence of several large peaks around position 895000 for the first three samples, and around position 952000 for the last two samples.
To ensure that our method performs sensibly, we pooled the reads for the GM12878 cell line and ran our method as well as a popular peak calling procedure MACS on the selected region. The results are shown in Figure \ref{fig:seq_peak}.\\
\newpage
\begin{figure}
\centering
    \begin{subfigure}[b]{0.85\textwidth}
        \centering
        \includegraphics[width=\textwidth]{peaks_comp_a.pdf}
        \caption{}
        \label{fig:seq_peak_data}
    \end{subfigure}
    \begin{subfigure}[b]{0.85\textwidth}
        \centering
        \includegraphics[width=\textwidth]{peaks_comp_b.pdf}
        \caption{}
        \label{fig:seq_peak_est}
    \end{subfigure}
    \caption{Analysis of reads taken from the GM12878 cell line, spanning positions 880001 to 1011072 on chromosome 1. (a) Raw sequencing counts. (b) Estimated intensity function from SMASH (black solid line) and location of peaks called by MACS (red markers beneath the estimated intensity).}
    \label{fig:seq_peak}
\end{figure}
From Figure \ref{fig:seq_peak_est} it is clear that SMASH can recognize the peaks detected by MACS. Rather than calling peaks based on certain thresholds however, our method allows users to determine the relative strength and width of each peak, which provides a more comprehensive summary of the sequencing reads. At the same time, SMASH also provides the posterior variances for the intensity estimates, allowing for the option of calling peaks based on thresholds if desired.\\

>>>>

We first apply the Gaussian version of SMASH on the well known motorcycle acceleration dataset from \cite{silverman1985some}. The data consists of 133 observations measuring head acceleration in a simulated motorcycle accident that is used to test crash helmets. The dependent variable is $acceleration$ (in $g$), and the independent variable is $time$ (in ms). To deal with repeated measurements, we simply take the median of the measurements for acceleration for any given $time$ value. To address the issue of non-equal spacing between the data points, we applied the procedure described in the Simulations section when comparing our method against MFVB. The fitted mean and variance curves are shown in Figure \ref{fig:motorcycle}. Comparing our fit with that of \cite{Delouille2004Smooth}, which also employs a wavelet-based approach, we note two advantages to using SMASH. First, our method provides a measure of uncertainty (through both the credible band for the mean curve as well as the variance curve itself), which is often of interest in many applications. Second, our method is able to produce a visually appealing fit when used without having to tune any parameters, while the approach in \cite{Delouille2004Smooth} requires an $adhoc$ removal of fine scale detail coefficients to avoid introducing artifacts into the fit.

Although SMASH does a reasonably good job of performing nonparametric regression on the given dataset, we do recognize that there are better ways to deal with the fact that the observations are not equally spaced, which can then reveal the discontinuities in the first derivative around 15, 23 and 32 ms, as discussed in the literature. Nevertheless, we can use SMASH directly out of the box in many settings where Gaussian denoising is required, making it an extremely versatile tool.
\newpage
\subsection{ChIP-Seq Data}
Here we apply our Poisson de-noising procedure to next generation sequencing data, commonly seen in the field of genomics. Specifically, we chose an example dataset from the ENCODE (\textbf{Enc}yclopedia \textbf{O}f \textbf{D}NA \textbf{E}lements) project launched by the National Human Genome Research Institute. This dataset contains reads from chromatin immunoprecipitation sequencing (ChIP-seq) measuring transcription factor binding in two different cell types, with two samples for each cell type. Due to the massive size of the data, we selected a representative portion of the reads of length $2^{15}$ from chromosome 1.
%Since we are looking mainly at univariate denoising here, we ran our method on each sample separately to produce the the plots in Figure \ref{fig:seq_smooth}.\\
%\begin{figure}[h]
%\centering
%\includegraphics[scale=0.7]{smoothing_all.pdf}
%\caption{Estimated intensities from SMASH for 4 samples; the first two samples are two replicates from GM12878, a lymphoblastoid cell line, and the last two samples are two replicates from H1hesc, which are human embryonic stem cells. The reads are taken from position 880001 to 1011072 on chromosome 1.}
%\label{fig:seq_smooth}
%\end{figure}
%\bigskip
One goal of analyzing ChIP-seq data is to discover regions where transcription factors are likely to bind to DNA, thereby allowing us to better understand the mechanisms underlying gene regulation. These binding regions are often reflected in the data as ``peaks'', where more counts are present than background noise. Our method allows us to identify these peaks by looking at the estimated intensity function.
%In Figure \ref{fig:seq_smooth}, we immediately notice the presence of several large peaks around position 895000 for the first three samples, and around position 952000 for the last two samples.
To ensure that our method performs sensibly, we pooled the reads for the GM12878 cell line and ran our method as well as a popular peak calling procedure MACS on the selected region. The results are shown in Figure \ref{fig:seq_peak}.\\
\newpage
\begin{figure}
\centering
    \begin{subfigure}[b]{0.85\textwidth}
        \centering
        \includegraphics[width=\textwidth]{peaks_comp_a.pdf}
        \caption{}
        \label{fig:seq_peak_data}
    \end{subfigure}
    \begin{subfigure}[b]{0.85\textwidth}
        \centering
        \includegraphics[width=\textwidth]{peaks_comp_b.pdf}
        \caption{}
        \label{fig:seq_peak_est}
    \end{subfigure}
    \caption{Analysis of reads taken from the GM12878 cell line, spanning positions 880001 to 1011072 on chromosome 1. (a) Raw sequencing counts. (b) Estimated intensity function from SMASH (black solid line) and location of peaks called by MACS (red markers beneath the estimated intensity).}
    \label{fig:seq_peak}
\end{figure}
From Figure \ref{fig:seq_peak_est} it is clear that SMASH can recognize the peaks detected by MACS. Rather than calling peaks based on certain thresholds however, our method allows users to determine the relative strength and width of each peak, which provides a more comprehensive summary of the sequencing reads. At the same time, SMASH also provides the posterior variances for the intensity estimates, allowing for the option of calling peaks based on thresholds if desired.\\
